plyranges: a grammar for transforming genomics data

### 2018-04-27 paper.tex
rebuild of paper

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{plyranges: a grammar for transforming genomics data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
Stuart Lee\textsuperscript{\getAff{Monash University}},
Michael Lawrence\textsuperscript{\getAff{Genentech}},
Di Cook\textsuperscript{\getAff{Monash University}}\\
\bigskip
\textbf{\getAff{Monash University}}Department of Econometrics and Business Statistics, Clayton, Victoria,
Australia\\
\textbf{\getAff{Genentech}}Bioinformatics and Computational Biology, Genentech, Inc., South San
Francisco, California, United States of America\\
\bigskip
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The Bioconductor project has created many useful data abstractions for
analysing high-throughput genomics experiments. However, there is a
cognitive load placed on a user in learning a data abstraction and
understanding its appropriate use. Throughout a standard workflow, a
user must navigate and know many of these abstractions to perform an
genomic analysis task, when a single data abstraction, a GRanges object
will suffice. The GRanges class naturally represent genomic intervals
and their associated measurements. By recognising that the GRanges class
follows `tidy' data principles we have created a grammar of genomic data
transformation. The grammar defines verbs for performing actions on and
between genomic interval data. It provides a principled way of
performing common genomic data analysis tasks through a coherent
interface to existing Bioconductor infrastructure, resulting in human
readable analysis workflows. We have implemented this grammar as an
Bioconductor/R package called plyranges.

@@

\section{Design and Implementation}\label{design-and-implementation}

The \texttt{plyranges} DSL is built on the most general Bioconductor
data structure, GRanges, which is capable of representing all types of
genomic data at a semantic level that roughly matches the intuition of
most users. A GRanges is essentially a table, with columns for the
chromosome, start and end coordinates, and the strand, along with an
arbitrary set of additional columns, consisting of measurements or
metadata specific to the data type or experiment. For example, a GRanges
can represent a gene with its chromosomal coordinates, exonic structure,
and a count column summarized from an RNA-seq experiment.

@@

\subsection{Genomic Relational
Algebra}\label{genomic-relational-algebra}

The \texttt{plyranges} DSL defines an expressive algebra for performing
genomic arithmetic with and between GRanges objects. Often in analyses
of genomic regions we are required to make adjustments to the start,
end, and width columns. This is supported directly in \texttt{plyranges}
via extending the \texttt{mutate()} function in \texttt{dplyr} (table
\ref{tab:grammar}). However, often adjustments need to be relative:
there is a mutual dependence between the start, end and width columns
over a genomic region. Changing the start,column needs to change either
the width or the end to preserve integrity of the object. The GRanges
object by design, causes a user to think of the range as a pair of
endpoints, and hence we have developed operators that perform coordinate
transformations with the notion of an `anchoring' combinator.

For width modification, we can stretch the range by a fixed amount or
resize the width of a range. The `anchoring' combinator preserves one of
the endpoints or the midpoint of a range when altering performing width
modification. It can also be used to preserve coordinates over strand,
for example, anchoring over the three prime end of a genomic region,
will preserve the start on the negative strand and end on the positive
strand. Anchoring the width would not make sense, because it is not a
point that can be intuitively anchored and stretching and resizing
obviously modify the width, so it would not apply to those.

Transformations that are width invariant are designed to be explicit
about the direction to modify and are performed via the shift operator.
For GRanges objects strand modulates direction and we distinguish
left/right (chromosomal coordinates) from the direction of
transcription. To generate flanking regions, we use the flank operator
which is also explicit about the direction of the input range.

Binary operations between two ranges mostly align with set operations
that treat ranges as sets of integers. These are parallel/vectorized
operations between two vectors/tables of ranges, and are not to be
confused with the merging operations defined in table \ref{tab:grammar}.
We make the binary nature of these operators by having in-fix functions
in R for the parallel union, intersect and asymmetric set difference.
There are also operations that relate to the relative positions of two
ranges such as the `between' operator which returns the range in the gap
between two ranges and the `span' operator a range that spans both
ranges. The binary operations can be expressed within the arithmetic,
restriction or aggregation functions in table \ref{tab:grammar}.

Our algebra recasts the actions of finding overlaps or nearest
neighbours between two genomic regions as a relational join operator.
The join operator acts on two GRanges object, a query and a subject. The
join operator is relational in the sense that metadata from the query
and subject ranges is retained in the joined range. The commonality
between all join operators in the \texttt{plyranges} DSL is to use the
index of where in the query range the subject range `hits' the query
range as a primary key (figure \ref{fig:olaps-fig}). The notion of
`hits' and hence the resulting key is determined by the type of join
(preceding, nearest, follows overlaps).

@@

We extend this idea by introducing three types of overlap joins: inner,
intersect and left (figure \ref{fig:olaps-fig}). The inner join
considers a hit to be any overlap between subject and query and expands
the query range to have length equal to the number of hits. The
intersect join uses the same definition of a hit as the inner join but
returns a range where the start and end coordinates are the intersect of
coordinates in the subject that hit the query. Finally, the overlap left
join is akin to left outer join in Cobb's relational algebra: it
performs an overlap inner join but also returns all query ranges that
are not hit by the subject.

The behaviour of a join operator can be further altered with additional
suffixes by restricting or expanding how a subject `hits' a query. For
example, for overlap joins we can use suffixes to encapsulate Allan's
Interval Algebra: to consider subject `hits' that are entirely `within'
a query range, the `within' prefix is used. We also use the `directed'
suffix to explicitly include notions of strandedness in our algebra.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{plyranges: a grammar for transforming genomics data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
Stuart Lee\textsuperscript{\getAff{Monash University}},
Michael Lawrence\textsuperscript{\getAff{Genentech}},
Di Cook\textsuperscript{\getAff{Monash University}}\\
\bigskip
\textbf{\getAff{Monash University}}Department of Econometrics and Business Statistics, Clayton, Victoria,
Australia\\
\textbf{\getAff{Genentech}}Bioinformatics and Computational Biology, Genentech, Inc., South San
Francisco, California, United States of America\\
\bigskip
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The Bioconductor project has created many useful data abstractions for
analysing high-throughput genomics experiments. However, there is a
cognitive load placed on a user in learning a data abstraction and
understanding its appropriate use. Throughout a standard workflow, a
user must navigate and know many of these abstractions to perform an
genomic analysis task, when a single data abstraction, a GRanges object
will suffice. The GRanges class naturally represent genomic intervals
and their associated measurements. By recognising that the GRanges class
follows `tidy' data principles we have created a grammar of genomic data
transformation. The grammar defines verbs for performing actions on and
between genomic interval data. It provides a principled way of
performing common genomic data analysis tasks through a coherent
interface to existing Bioconductor infrastructure, resulting in human
readable analysis workflows. We have implemented this grammar as a
Bioconductor/R package called plyranges.

@@

\section{Genomic relational algebra}\label{genomic-relational-algebra}

\subsection{Data model}\label{data-model}

The \texttt{plyranges} DSL is built on the core Bioconductor data
structure GRanges, which is capable of representing all types of genomic
data at a semantic level that roughly matches the intuition of most
users. A GRanges is essentially a table, with columns for the
chromosome, start and end coordinates, and the strand, along with an
arbitrary set of additional columns, consisting of measurements or
metadata specific to the data type or experiment. For example, a GRanges
can represent a gene with its chromosomal coordinates, exonic structure,
and a count column summarized from an RNA-seq experiment.

@@

\subsection{Algebraic operations}\label{algebraic-operations}

The \texttt{plyranges} DSL defines an expressive algebra for performing
genomic operations with and between GRanges objects (see table
\ref{tab:grammar}). {[}ML: add a couple overview sentences indicating
the classes of operations like arithmetic, merging, aggregation{]}

GRanges describes the within-sequence coordinates of a range by its
\emph{start}, \emph{end} and \emph{width}. Those three variables are
mutually dependent and partially redundant, so direct manipulation of
them is problematic. For example, changing the \emph{width} column needs
to change either the \emph{start}, \emph{end} or both to preserve
integrity of the object. We introduce the \emph{anchor} combinator to
disambiguate these adjustments. Supported anchor points include the
start, end and midpoint, as well as the 3' and 5' ends for
strand-directed ranges. For example, if we anchor the start, then
setting the width will adjust the end while leaving the start
stationary.

{[}ML: where does the word `combinator' come from here? maybe `modifier'
is better? that's what the table uses{]}

The algebra also defines conveniences for relative coordinate
adjustments: \emph{shift} (unanchored adjustment to both start and end)
and \emph{stretch} (anchored adjustment of width). We can perform any
relative adjustment by some combination of those two operations. The
\emph{stretch} operation requires an anchor and assumes the midpoint by
default. Since \emph{shift} is unanchored, the user specifies a suffix
for indicating the direction: left/right or, for stranded features,
upstream/downstream. For example, \emph{shift\_right} shifts a range to
the right.

The \emph{flank} operation generates new ranges that are adjacent to
existing ones. This is useful, for example, when generating upstream
promoter regions for genes. Analogous to \emph{shift}, a suffix
indicates the side of the input range to flank.

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including \emph{intersect}, \emph{union},
\emph{difference}, and \emph{complement}. There are two sets of these:
parallel and merging. The parallel operations map to infix operators,
which we surround with \emph{\%} symbols in accordance with R syntax
rules. For example, the parallel intersection (\emph{x \%intersect\% y})
finds the intersecting range between \emph{xi} and \emph{yi} for
\emph{i} in \emph{1\ldots{}n}, where \emph{n} is the length of both
\emph{x} and \emph{y}. In constrast, the merging intersection
(\emph{intersect\_ranges(x, y)}) returns a new set of disjoint ranges
representing whereever there was overlap between a range in \emph{x} and
a range in \emph{y}. We use the infix syntax for the parallel
operations, since it is the conventional syntax for parallel, binary
operations in R. Finding the parallel union will fail when two ranges
have a gap, so we introduce a \emph{span} operator that takes the union
while filling any gap. The \emph{complement} operation is unique in that
it is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the \emph{between} parallel operation,
which finds the gap separating \emph{xi} and \emph{yi}. The binary
operations are callable from within the arithmetic, restriction and
aggregation expressions.

Our algebra recasts finding overlaps or nearest neighbours between two
genomic regions as variants of the relational join operator. A join acts
on two GRanges objects, a query and a subject. The join operator is
relational in the sense that metadata from the query and subject ranges
is retained in the joined range. All join operators in the
\texttt{plyranges} DSL generate a set of hits based on overlap or
proximity of ranges and use those hits to merge the two datasets in
different ways. There are four supported matching algorithms:
\emph{overlap}, \emph{nearest}, \emph{precede}, and \emph{follow}. We
can further restrict the matching by whether the query is completely
\emph{within} the subject, and adding the \emph{directed} suffix ensures
that matching ranges have the same direction (strand).

For merging based on the hits, we have three modes: \emph{inner},
\emph{intersect} and \emph{left}. The \emph{inner} overlap join is
similar to the conventional inner join in that there is a row in the
result for every match. A major difference is that the matching is not
by identity, so we have to choose one of the ranges from each pair. We
always choose the left range. The \emph{intersect} join uses the
intersection instead of the left range. Finally, the overlap \emph{left}
join is akin to left outer join in Cobb's relational algebra: it
performs an overlap inner join but also returns all query ranges that
are not hit by the subject.

@@

{[} add a simple paragraph about how we can filter, sort, group and
aggregate just like other tabular structures {]}
